Pre-processing high frequency animal tracking data

### 2020-09-17 02_filtering_data.Rmd
updates

```{r read_sim_data}
# read in the data
data <- fread("data/data_sim.csv")
```

@@

```{r add_outlier}
# add 100 outliers
data_copy <- do_add_outliers(data, p_data = 0.01, std_dev = 0.05)
```

## Removing gross positioning errors

@@

```{r remove_outside_bbox, eval=FALSE}
# remove positions outside a bounding box
data_inside_bbox <- data.table::copy(data_copy)

# NB: set remove_inside to FALSE
data_inside_bbox <- atlastools::atl_filter_bounds(data = data,
                                                  x_range = c(0, 1),
                                                  y_range = c(0, 1),
                                                  remove_inside = FALSE)
```

```{r}
ggplot()+
  geom_path(data = data_copy, aes(x, y), col = "red")+
  geom_path(data = data_inside_bbox, aes(x, y))
```

@@

```{r example_remove_outliers, eval=FALSE}
# get speed and turning angle
data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs,
                                              type = "in"),
                     out_speed = atl_get_speed(data_many_nbs,
                                               type = "out"),
                     angle = atl_turning_angle(data_many_nbs))]

@@

```{r}
# make a copy of the data
data_non_move <- copy(data_many_nbs)

# filter the copy
data_non_move <- atl_filter_covariates(data_non_move,
                                       filters = c("in_speed < 0.001",
                                                   "out_speed < 0.001",
                                                   "angle < 40"))
```

```{r}
ggplot()+
  geom_path(data = data_inside_bbox,
            aes(x, y),
            colour = "black")+
  geom_path(data = data_non_move,
            aes(x, y),
            colour = "red")
```

@@

```{r outlier_error_combinations}
# define unique values
p_outlier <- c(0.05)
ext_outlier <- c(0.01, 0.05, 0.1)
land_scale <- c(1, 2, 5, 10)
replicate <- seq_len(10)

# make combinations
param_c <- CJ(p_outlier, ext_outlier, land_scale, replicate)
```

### Procedural landscape generation

Next we procedurally generate (simulate) a landscape from which to extract habitat choice.

```{r make_landscape}
# use NLMR
landscape <- NLMR::nlm_randomcluster(100, 100, 
                                     p = 0.2, 
                                     ai = c(0.2, 0.3, 0.5),
                                     rescale = TRUE)

# save landscape as raster
raster::writeRaster(landscape,
                    filename = "data/data_landscape.tif",
                    format = "GTiff", 
                    overwrite = TRUE)
```

```{r read_landscape}
# read landscape
landscape <- raster("data/data_landscape.tif")
```

### Simulate movement over a landscape

```{r}
sim_data <- mapply(function(p_outlier, ext_outlier, scale) {

  # scale the landscape
  tmp_land <- do_scale_landscape(land_ = landscape,
                                 scale_factor = scale)

  # make simulated data
  ref_data <- do_smoove_data()

  # extract 'true' movement metrics
  speed <- atl_get_speed(ref_data, type = "in")
  speed
  land_vals <- do_count_landcover(tmp_land, ref_data)
  land_vals[, type := "reference"]

  # now add error
  error_data <- do_add_outliers(ref_data, 
                                p_data = p_outlier,
                                std_dev = ext_outlier)

  # get movement metrics with error
  error_speed <- atl_get_speed(error_data, type = "in")
  error_land_vals <- do_count_landcover(tmp_land, error_data)
  error_land_vals[, type := "outliers"]

  # now clean data
  clean_data <- copy(error_data)
  clean_data[, `:=`(in_speed = atl_get_speed(clean_data, type = "in"),
                    out_speed = atl_get_speed(clean_data, type = "out"),
                    angle = atl_turning_angle(clean_data))]
  upper_speed_cutoff <- stats::quantile(clean_data$in_speed, 
                                        probs = 0.9,
                                        na.rm = TRUE)
  angle_cutoff <- 40
  clean_data <- atl_filter_covariates(clean_data,
                                      filters = c(glue('{c("in_speed", 
                                                           "out_speed")} < 
                                                         {upper_speed_cutoff}'),
                                                  glue('angle < {angle_cutoff}')
                                                  )
                                      )

  # get speed and land proportion
  clean_speed <- atl_get_speed(clean_data, type = "in")
  clean_land_vals <- do_count_landcover(tmp_land, clean_data)
  clean_land_vals[, type := "cleaned"]

  # synthesize data
  land_props <- rbindlist(list(land_vals, error_land_vals, clean_land_vals))
  # count proportion
  land_props <- dcast(land_props, land_values ~ type, value.var = "p_land")

  land_props[, `:=`(scale = scale,
                    p_outlier = p_outlier,
                    ext_outlier = ext_outlier)]

  # make speed results
  speed_res <- data.table(type = c("reference", "outliers", "cleaned"),
                          speed_mean = unlist(lapply(list(speed, 
                                                          error_speed, 
                                                          clean_speed),
                                                     mean, na.rm = TRUE)),
                          speed_sd = unlist(lapply(list(speed, 
                                                          error_speed, 
                                                          clean_speed),
                                                     sd, na.rm = TRUE)),
                          scale = scale,
                          p_outlier = p_outlier,
                          ext_outlier = ext_outlier
                          )
  results <- list(land_props = land_props,
                  speed_res = speed_res)

  return(results)
}, 
param_c$p_outlier, 
param_c$ext_outlier,
param_c$land_scale,
SIMPLIFY = FALSE)
```

```{r outlier_removal_land_prop}
# get bound list
land_prop_data <- rbindlist(lapply(sim_data, function(l) { l$land_props }))
# get differences between outliers and cleaned wrt ref
#land_prop_data[, `:=`(d_cleaned = abs(cleaned - reference),
 #               d_outliers = abs(outliers - reference))]

land_prop_data_long <- copy(land_prop_data)

land_prop_data_long <- melt(land_prop_data_long, 
                            id.vars = c("land_values", "ext_outlier", 
                                        "p_outlier", "scale"))

ggplot(land_prop_data_long)+
  # geom_boxplot(aes(x = factor(land_values),
  #                  y = value,
  #                  group = interaction(variable, scale, land_values),
  #                  fill = variable))+
  geom_line(aes(x = (land_values),
                y = value,
                group = interaction(variable, scale),
                col = variable,
                lty = factor(scale)),
            stat = "summary")+
  geom_point(aes(x = (land_values),
                 y = value,
                 group = interaction(variable, scale),
                 col = variable,
                 shape = factor(scale)),
             size = 2,
             stat = "summary")+
  facet_grid( ~ ext_outlier, 
              scales = "free_y",
              labeller = label_both)+
  theme_classic()+
  theme(strip.background = element_rect(fill = "grey90",
                                        colour = NA))
```

```{r outlier_removal_speed}
# get speed data
speed_data <- rbindlist(lapply(sim_data, function(l) { l$speed_res }))

# melt to
speed_data_long <- copy(speed_data)

speed_data_long <- melt(speed_data_long, 
                            id.vars = c("ext_outlier", "type",
                                        "p_outlier", "scale"))

ggplot(speed_data_long)+
  # geom_boxplot(aes(x = factor(land_values),
  #                  y = value,
  #                  group = interaction(variable, scale, land_values),
  #                  fill = variable))+
  geom_boxplot(aes(x = factor(ext_outlier),
                   y = value,
                   col = type))+
  facet_grid(~ variable, 
              scales = "free_y",
              labeller = label_both)+
  theme_classic()+
  theme(strip.background = element_rect(fill = "grey90",
                                        colour = NA))
```

@@

```{r}
# make a copy of data with spikes removed
data_with_reflection <- copy(data_non_move)

@@

@@

```{r}
# attempt to remove reflections
data_no_reflection <- copy(data_with_reflection)

data_no_reflection <- atl_remove_reflections(data_with_reflection,
                                              point_angle_cutoff = 10,
                                              reflection_speed_cutoff = 0.0015)
```

```{r}
ggplot()+
  geom_path(data = data_with_reflection,
            aes(x, y))+
  geom_path(data = data_no_reflection,
            aes(x, y),
            colour = "red")
```

## Assessing whether filtering helps

>>>>

```{r read_sim_data}
# read in the data
data <- fread("data/data_sim.csv")[5000:10000, ]
```

@@

```{r add_outlier}
# make a dopy
data_copy <- copy(data)

# add normal error
data_copy[, `:=`(x = do_add_error(x, std_dev = 0.01),
                 y = do_add_error(y, std_dev = 0.005))]
# add 100 outliers
data_copy <- do_add_outliers(data_copy, p_data = 0.005, std_dev = 0.1)
```

## Show data with errors

```{r}
# make list of data to plot
data_plot <- list(data, data_copy)

# make plot list
plots <- lapply(data_plot, function(df) {
  ggplot(data = df)+

    geom_path(aes(x, y),
              col = "red", 
              alpha = 1, lwd = 0.2)+
    geom_point(aes(x, y),
              col = "darkblue", 
              alpha = 0.1,
              shape = 16)+
    # coord_cartesian(xlim = c(0.5, 1),
    #                 ylim = c(0.5, 1)) +
    coord_equal()+
    theme_test()+
    theme(axis.text = element_blank(),
          axis.title = element_blank())
})

# wrap plots
plot_figure <- wrap_plots(plots) +
  plot_annotation(tag_levels = "a",
                  tag_prefix = "(",
                  tag_suffix = ")") &
  theme(plot.tag = element_text(face = "bold"))

plot_figure

# save figure
ggsave(plot_figure, filename = "figures/fig_add_errors.png",
       width = 185, height = 90, units = "mm")
```

## Filtering by spatial bounds

@@

```{r remove_outside_bbox, eval=FALSE}
# remove positions outside a bounding box
# NB: set remove_inside to FALSE
data_inside_bbox <- atlastools::atl_filter_bounds(data = data_copy,
                                                  y_range = c(0.5, 1),
                                                  remove_inside = FALSE)
```

```{r}
# data to plot
data_plot <- list(data_copy, data_inside_bbox)

# make plots
plots <- lapply(data_plot, function(df) {
  ggplot()+

    geom_path(data = df,
              aes(x, y),
              col = "red", 
              alpha = 1, lwd = 0.2)+
    geom_point(data = df,
               aes(x, y),
              col = "darkblue", 
              alpha = 0.1,
              pch = 16)+
    theme_test()+
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank())
})

# add elements
plots[[1]] <- plots[[1]] + 
  annotate(xmin = 0.5, xmax = 1,
           ymin = 0.5, ymax = 1,
           fill = NA, col = "black", 
           lty = 2, geom = "rect")+
  annotate(x = 0.5, y = 1.1, label = "(b)", 
           fontface = "bold", geom = "text", size = 4)

# wrap plots
plot_figure <- wrap_plots(plots) +
  plot_annotation(tag_levels = "a",
                  tag_prefix = "(",
                  tag_suffix = ")") &
  theme(plot.tag = element_text(face = "bold"))

plot_figure

ggsave(plot_figure, filename = "figures/fig_filter_bounds.png", 
       width = 185, height = 90, units = "mm")
```

@@

```{r example_remove_outliers, eval=FALSE}
# get speed and turning angle
data_copy[, `:=`(in_speed = atl_get_speed(data_copy,
                                          type = "in"),
                 out_speed = atl_get_speed(data_copy,
                                           type = "out"),
                 angle = atl_turning_angle(data_copy))]

@@

```{r}
# filter the copy by the 95th percentile
data_filtered <- atl_filter_covariates(data_copy,
            filters = c("(in_speed < 0.03 & out_speed < 0.03) | angle < 40"))
```

```{r}
# data plot
data_plot <- list(data_copy, data_filtered)

plots <- lapply(data_plot, function(df) {
  ggplot()+
     geom_path(data = df,
              aes(x, y),
              col = "red", 
              alpha = 1, lwd = 0.2)+
    geom_point(data = df,
               alpha = 0.1,
               aes(x, y, 
                   col = (in_speed >= 0.03 & out_speed >= 0.03)),
               shape = ifelse((df$in_speed >= 0.03 & df$out_speed >= 0.03), 
                              16, 1),
               show.legend = F)+
    scale_colour_manual(values = c("darkblue", "red"))+
    theme_test()+
    coord_equal()+
    theme(axis.title = element_blank(),
          axis.text = element_blank())
})

plot_figure <- wrap_plots(plots) +
  plot_annotation(tag_levels = "a",
                  tag_prefix = "(",
                  tag_suffix = ")") &
  theme(plot.tag = element_text(face = "bold"))

plot_figure
ggsave(plot_figure, filename = "figures/fig_filter_spikes.png", 
       width = 185, height = 90, units = "mm")
```

@@

```{r outlier_error_combinations}
# define unique values
p_outlier <- c(0.05)
ext_outlier <- c(0.01, 0.05, 0.1)
replicate <- seq_len(10)

# make combinations
param_c <- CJ(p_outlier, ext_outlier, replicate)
```

@@

```{r}
# make a copy of data with spikes removed
data_with_reflection <- copy(data_filtered)

@@

In some instances, the estimated location is instantaneously displaced from the real movement track for multiple consecutive timesteps in the sampling process, followed by a return to the real position.
Removing such prolonged spikes (which we call `reflections') using Bjorneraas and colleagues' approach (2010) would be ineffective --- the bounds of the reflection have either extreme incoming \textit{or} outgoing speeds, and there are no positions with both.
When encountering such data, users are advised to identify the bounds of the reflection as first step to removing them: the inner bound (where the position estimate first shifts away from the track) will have a high incoming speed ($v_i$), while the outer bound (after which the position estimate returns to the real track) will have a high outgoing speed ($v_o$).
We suggest locating the first reflected point $p_1$ by comparing the incoming speeds of the track positions ($v_i$) against the speed cutoff $S$.
An optional turning angle cutoff $A$ may be used if the track includes plausible fast transit segments.
The first position to satisfy this comparison $v_i > S$ AND $\theta > A$ can be considered the inner bound of the reflection and set as an `anchor point'.
The position after the anchor point with the highest outgoing speed $v_o$ (such that $v_o > S$) can be considered to be the outer bound of the reflection.
The positions between these bounds can be identified as a trajectory reflection and removed.
For a system-specific reasons and the pre-processing steps thus far, a reflection may not end before the last position in the movement track.
In such cases, the entirety of the track following the first anchor point should be removed, as it is better to exclude moderate chunks of data than to make inferences from unsound data.
Removing reflections is implemented in a convenient form in \texttt{atlastools} in the \texttt{{atl\_remove\_reflections}} function.

@@

```{r}
# attempt to remove reflections
data_no_reflection <- atl_remove_reflections(data_with_reflection,
                                              point_angle_cutoff = 10,
                                              reflection_speed_cutoff = 0.03)
```

```{r}
# data to plot
plot_data <- list(data_with_reflection, data_no_reflection)

# get plots
plots <- lapply(plot_data, function(df) {
  ggplot()+
    geom_path(data = df,
              aes(x, y),
              col = "red", 
              alpha = 1, lwd = 0.2)+
    geom_point(data = df,
               aes(x, y),
               alpha = 0.1,
               colour = "darkblue",
               shape = 16,
               show.legend = F)+
    scale_colour_manual(values = c("darkblue", "red"))+
    theme_test()+
    coord_equal()+
    theme(axis.title = element_blank(),
          axis.text = element_blank())
})

# wrap figures
plot_figure <- wrap_plots(plots) +
  plot_annotation(tag_levels = "a",
                  tag_prefix = "(",
                  tag_suffix = ")") &
  theme(plot.tag = element_text(face = "bold"))

plot_figure

ggsave(plot_figure, filename = "figures/fig_filter_reflection.png", 
       width = 185, height = 90, units = "mm")
```

## Export data

```{r}
fwrite(data_no_reflection, file = "data/data_no_reflection.csv")
```
